Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-08-23 article/article.Rmd
Commit 2024-08-23 20:14

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OIR). The OIR is defined as the irradiance difference of the measured
one-minute GHI from the $\text{GHI}_\text{ref}$ corresponding to cloud-free
atmosphere.
<!-- CE identification criterion ($\text{OIR}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023].
<!--
, and the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.
-->
We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the actual CE events.  However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric conditions
introduced numerous false positive or false negative results. The main reason for
these discrepancies is the variability of the effects of aerosols and water vapor
which were not taken into account in the two simple methods.  To produce a more
representative reference we included the effects of these factors using a radiative
transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}.

@@

In this study our main focus was to quantify the over irradiance related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the clear-sky irradiance at the time of each GHI
measurement. This depends on the selection of the appropriate atmospheric
parameterization for the RTM simulations. The implementation of the long-term change
in AOD, discussed in section\nobreakspace{}\ref{rtmcs}, allows capturing a large part
of the natural variability of clear-sky GHI. However, the short-term variability of
AOD cannot be taken adequately into account when using monthly values in the model
simulations. We tried different approaches in order to strengthen the robustness of
the methodology and to compensate for the limited accuracy of the RTM input data and
the unpredictable natural variability of the atmosphere.

@@

This is the criterion of our CE identification.  The constant terms were determined
through the implementation of an empirical method with manual inspection of the CE
identification results on selected days of the whole dataset.  We tested seven sets
of days with different characteristics relevant to the efficiency of the
identification threshold.  These sets were random groups of about 20 -- 30 days with
the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of CE events,
(e) with at least $60\,\%$ of the day length without clouds and presence of CE events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

<!-- reasoning for criteria -->
<!-- We have to note, that the differentiation of the threshold factor was needed, because
<!-- of the high irradiance values we observed, early in the morning and late in the
<!-- afternoon. 
<!-- We have confirmed, by inspecting images from the sky cam, that we have
<!-- duration of elevated irradiance, either due to the clearness of the atmosphere, or some
<!-- interferences by reflections on nearby bright surfaces. Although, these cases may
<!-- mask some of the CE events. The actual CE events, produce higher irradiance and thus
<!-- are identified as such. As a side effect, the reported OIR will be slight
<!-- underestimated for those SZAs, but the overall contribution of those cases to the
<!-- total daily energy, is minimal due the low occurrences and the lower irradiances.
-->

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method was proven capable in detecting all major CE events.  Where some CE
events with very low OIR may be not detected, these are few with small
over-irradiance, and it is unlikely that will affect significantly our results.

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the relevant literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold $E$ is given by:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: E > \cos(\theta) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for
the actual Sun -- Earth distance $r$ and $r_\text{m}$ is the mean Sun -- Earth
distance of $1.496\times10^8\,\text{km}$.

@@

Following the application of the above discussed methodology to the entire dataset 
($`r round(ST_total$wattGLB.N / 10^6,1)`$ million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OIR of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OIR of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The strongest CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OIR and excess irradiation.

@@

One aspect of this study is to investigate the time evolution of the CE events by
analyzing the GHI measurements at Thessaloniki. Cloud enhancements can be influenced
by different factors, such as the geometry, size and optical thickness of clouds,
their height in the atmosphere and local weather regimes [@Mol2023; @Veerman2022;
@Gristey2022; @Tzoumanikas2016].  Some of these factors are related to changes in
climate; hence it would be reasonable to expect contributing to the frequency of
occurrence of CE events over Thessaloniki, as well as to the average OIR and excess
irradiation.  The long-term trends were calculated using a first-order autoregressive
model with the 'maximum likelihood' fitting method [@Gardner1980; @Jones1980], by
implementing the function 'arima' from the library 'stats' of the R programming
language [@RCT2023]. All trends are reported together with their $2\sigma$ error.

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OIR and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OIR is negligible 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OIR for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1993 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OIR and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

The distribution of the number of CE events as a function of OIR is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}.  There is an inverse
relationship between the frequency of CE events and OIR with an exponential-like
decline. This is expected, as the stronger the CE events are, the rarer the
conditions favoring the occurrence of CE events.  For the majority (over $62\,\%$) of
the CE events the OIR is below the long-term average of
$`r round(mean(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$, while about
$8.1\,\%$ of the events correspond to OIR larger than $100\,\text{W}/\text{m}^2$ and
up to the highest value of $412.4\,\text{W}/\text{m}^2$.  This distribution is
indicative of the magnitude and the probability of the expected CE events over
Thessaloniki.  Similar distribution of CE events, albeit with larger OIR values, has
been reported by @Vamvakas2020, for the city of Patras. This site is located
$2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly from the
eastern Mediterranean resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OIR."}
knitr::include_graphics("../images/P-relative-distribution-diff-1.png")
```

@@

The relation between the duration and the mean OIR of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, events of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g., $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g., $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been reported
by @Zhang2018 in a study using a far higher sampling rate ($100\,\text{Hz}$) than
ours.

@@

<!-- ```{r echo=FALSE, fig.cap="Relation of mean over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-4.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Relation of maximun over irradiance and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-8.png") -->
<!-- ``` -->

@@

An aspect of the CE events that is commonly reported and has some significance on the
solar energy production infrastructure are the extreme CE events (ECE), where solar
irradiance exceeds the expected irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}).  Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have computed the
distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in spring and early summer (March -- June), followed by a
period in late autumn (September and October).  This is probably related to the
weather characteristics in these periods, with frequent alternations between clear
sky periods and clouds.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in rare cases the
OIR can exceed the TSI even by more than $400\,\text{W}/\text{m}^2$, while in
$75\,\%$ of the cases the OIR is below $200\,\text{W}/\text{m}^2$. The OIR for the
most frequent ECEs ranges between $140$ and $180\,\text{W}/\text{m}^2$.  These
findings are in accordance with the results of @Vamvakas2020, the only difference
being that the OIR values reported for are higher than those for Thessaloniki.

@@

We have observed ECE cases exceeding the TSI by $400\,\text{W}/\text{m}^{2}$ with
$75\,\%$ of the cases under $200\,\text{W}/\text{m}^{2}$. The climatological
characteristics of the ECE events showed that the most active months are spread in
half of the year and particularly in the periods March -- June and September --
October.  The magnitude of the ECE events identified in Thessaloniki events does not
exceed the values reported for sites with more favourable conditions for the
phenomenon [e.g., @Cordero2023]. Some of the characteristics of CE and ECE events we
analyzed have strong similarities with the results of @Vamvakas2020 for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude of OIR.

>>>>

For the detection of CE cases we established a baseline of irradiance above which we
characterized each data point as an enhancement event and calculated the over
irradiance (OI). The OI is defined as the irradiance difference of the measured
one-minute GHI from the $\text{GHI}_\text{ref}$ corresponding to cloud-free
atmosphere.
<!-- CE identification criterion ($\text{OI}_i =
\text{GHI}_i - \text{GHI}_\text{CSlim}$), as defined in
Equation\nobreakspace\ref{eq:CE4}. -->
First, we tested two simple approaches for the determination of
$\text{GHI}_\text{ref}$: The Haurwitz's model [@Haurwitz1945], which is a simple
clear sky model and was already adjusted and applied to our data [@Natsis2023].
<!--
, and the total solar irradiance (TSI) at the top of the atmosphere, adjusted for the
Sun-Earth distance.
-->
We have tested both methods by using an appropriate relative
threshold and/or an additional constant offset.  The initial results showed that we
can detect a big portion of the actual CE events.  However, by inspecting the daily
plots of irradiance it became evident that changes in atmospheric conditions
introduced numerous false positive or false negative results. The main reason for
these discrepancies is the variability of the effects of aerosols and water vapor
which were not taken into account in the two simple methods.  To produce a more
representative reference we included the effects of these factors using a radiative
transfer model (RTM).  The applied methodology is discussed in
section\nobreakspace{}\ref{rtmcs}.

@@

In this study our main focus was to quantify the OI related to CEs. A
key issue for achieving this goal is to define a threshold for the CE
identification, representative of the clear-sky irradiance at the time of each GHI
measurement. This depends on the selection of the appropriate atmospheric
parameterization for the RTM simulations. The implementation of the long-term change
in AOD, discussed in section\nobreakspace{}\ref{rtmcs}, allows capturing a large part
of the natural variability of clear-sky GHI. However, the short-term variability of
AOD cannot be taken adequately into account when using monthly values in the model
simulations. We tried different approaches in order to strengthen the robustness of
the methodology and to compensate for the limited accuracy of the RTM input data and
the unpredictable natural variability of the atmosphere.

@@

This is the criterion of our CE identification.  The constant terms were determined
through the implementation of an empirical method with manual inspection of the CE
identification results on selected days of the whole dataset.  We tested seven sets
of days with different characteristics relevant to the efficiency of the
identification threshold.  These sets were random groups of about 20 -- 30 days with
the following characteristics:
(a) the strongest OI CE events,
(b) the largest daily total OI,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of CE events,
(e) with at least $60\,\%$ of the day length without clouds and presence of CE events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

<!-- reasoning for criteria -->
<!-- We have to note, that the differentiation of the threshold factor was needed, because
<!-- of the high irradiance values we observed, early in the morning and late in the
<!-- afternoon. 
<!-- We have confirmed, by inspecting images from the sky cam, that we have
<!-- duration of elevated irradiance, either due to the clearness of the atmosphere, or some
<!-- interferences by reflections on nearby bright surfaces. Although, these cases may
<!-- mask some of the CE events. The actual CE events, produce higher irradiance and thus
<!-- are identified as such. As a side effect, the reported OI will be slight
<!-- underestimated for those SZAs, but the overall contribution of those cases to the
<!-- total daily energy, is minimal due the low occurrences and the lower irradiances.
-->

@@

The definition of the CE events with this method has a degree of subjectivity, since
the actual clear sky irradiance is not known and can only be approximated.  However,
this method was proven capable in detecting all major CE events.  Where some CE
events with very low OI may be not detected, these are few with small
over-irradiance, and it is unlikely that will affect significantly our results.

<!-- extreme caces criterion -->
A sub-category of the CE events that is often discussed in the literature
[@Cordero2023; @Martins2022; @Yordanov2015], are the extreme cloud enhancement (ECE)
events. These are cases of CE where the measured intensity of the irradiance exceeds
the TSI at the top of the atmosphere. In this case the threshold $E$ is given by:
<!--and detected according to the Equation\nobreakspace{}\ref{eq:ECE}.-->
\begin{equation}
\text{ECE}: E > \cos(\theta) \cdot E_{\odot} \frac{r^2_\text{m}} {r^2}
\label{eq:ECE}
\end{equation}
where: $\theta$ the solar zenith angle, $E_{\odot}$ the solar constant adjusted for
the actual Sun -- Earth distance $r$ and $r_\text{m}$ is the mean Sun -- Earth
distance of $1.496\times10^8\,\text{km}$.

@@

Following the application of the above discussed methodology to the entire dataset 
($`r round(ST_total$wattGLB.N / 10^6,1)`$ million of one-minute GHI measurements),
$`r round(100*ST_total$GLB_diff.N_pos/ST_total$wattGLB.N, 2)`\,\%$ were identified as CE events and
$`r round(100*ST_extreme_total$GLB_diff.N_pos/ST_total$wattGLB.N, 3)`\,\%$ as ECE events.
The highest recorded GHI due to CE was
$`r round(ST_daily[which.max(wattGLB.max), wattGLB.max],1)`\,\text{W}/\text{m}^2$
on `r strftime(ST_daily[which.max(wattGLB.max), Date], "%d\\ %B %Y")`, corresponding
to OI of $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 1)`\,\text{W}/\text{m}^2$.
<!-- The absolute strongest CE event had an OI of 
<!-- $`r round(ST_daily[which.max(GLB_diff.max), GLB_diff.max], 2)`\,\text{W}/\text{m}^2$ on
<!-- `r strftime(ST_daily[which.max(GLB_diff.max), Date], "%d\\ %B %Y")`. -->
The strongest CE event of
$`r round(100*ST_daily[which.max(GLB_ench.max), GLB_ench.max],1)`\,\%$ above the 
clear sky threshold was observed on 
`r strftime(ST_daily[which.max(GLB_ench.max), Date], "%d\\ %B %Y")`.
In the following sections we are discussing the long-term trends and variability of
the CE events as well as of the corresponding OI and excess irradiation.

@@

One aspect of this study is to investigate the time evolution of the CE events by
analyzing the GHI measurements at Thessaloniki. Cloud enhancements can be influenced
by different factors, such as the geometry, size and optical thickness of clouds,
their height in the atmosphere and local weather regimes [@Mol2023; @Veerman2022;
@Gristey2022; @Tzoumanikas2016].  Some of these factors are related to changes in
climate; hence it would be reasonable to expect contributing to the frequency of
occurrence of CE events over Thessaloniki, as well as to the average OI and excess
irradiation.  The long-term trends were calculated using a first-order autoregressive
model with the 'maximum likelihood' fitting method [@Gardner1980; @Jones1980], by
implementing the function 'arima' from the library 'stats' of the R programming
language [@RCT2023]. All trends are reported together with their $2\sigma$ error.

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OI and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OI is negligible 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OI for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\begin{figure}% [h!]
        {\centering 
            \subfloat[\label{fig:P-energy-mean}]
                {\includegraphics[width=\linewidth]{../images/P-energy-3} }\\
            \subfloat[\label{fig:P-energy-N}]
                {\includegraphics[width=\linewidth]{../images/P-energy-2} }\\
            \subfloat[\label{fig:P-energy-sum}]
                {\includegraphics[width=\linewidth]{../images/P-energy-1} }
        }
    \caption{Time series for the period 1993 -- 2023 of (a) the yearly CE number of occurrences, (b) the yearly mean OI and (c) the yearly excess irradiation. The black lines represent the linear trends on the yearly data.}\label{fig:P-energy}
\end{figure}

@@

The distribution of the number of CE events as a function of OI is shown in
Figure\nobreakspace{}\ref{fig:ovir-distribution}.  There is an inverse
relationship between the frequency of CE events and OI with an exponential-like
decline. This is expected, as the stronger the CE events are, the rarer the
conditions favoring the occurrence of CE events.  For the majority (over $62\,\%$) of
the CE events the OI is below the long-term average of
$`r round(mean(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$, while about
$8.1\,\%$ of the events correspond to OI larger than $100\,\text{W}/\text{m}^2$ and
up to the highest value of $412.4\,\text{W}/\text{m}^2$.  This distribution is
indicative of the magnitude and the probability of the expected CE events over
Thessaloniki.  Similar distribution of CE events, albeit with larger OI values, has
been reported by @Vamvakas2020, for the city of Patras. This site is located
$2.5^\circ$ south of Thessaloniki and is exposed to air masses coming mainly from the
eastern Mediterranean resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

```{r ovir-distribution, echo=FALSE, fig.cap="Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OI."}
knitr::include_graphics("../images/P-relative-distribution-diff-1.png")
```

@@

The relation between the duration and the mean OI of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, events of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g., $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g., $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been reported
by @Zhang2018 in a study using a far higher sampling rate ($100\,\text{Hz}$) than
ours.

@@

<!-- ```{r echo=FALSE, fig.cap="Relation of mean OI and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-4.png") -->
<!-- ``` -->

<!-- ```{r echo=FALSE, fig.cap="Relation of maximun OI and CE group duration"} -->
<!-- knitr::include_graphics("../images/groups-8.png") -->
<!-- ``` -->

@@

An aspect of the CE events that is commonly reported and has some significance on the
solar energy production infrastructure are the extreme CE events (ECE), where solar
irradiance exceeds the expected irradiance on top of the atmosphere at the same SZA
(Equation\nobreakspace{}\ref{eq:ECE}).  Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} we have computed the
distribution of the number of occurrences of ECE events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most active period
for ECE events is in spring and early summer (March -- June), followed by a
period in late autumn (September and October).  This is probably related to the
weather characteristics in these periods, with frequent alternations between clear
sky periods and clouds.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}) shows that in rare cases the
OI can exceed the TSI even by more than $400\,\text{W}/\text{m}^2$, while in
$75\,\%$ of the cases the OI is below $200\,\text{W}/\text{m}^2$. The OI for the
most frequent ECEs ranges between $140$ and $180\,\text{W}/\text{m}^2$.  These
findings are in accordance with the results of @Vamvakas2020, the only difference
being that the OI values reported for are higher than those for Thessaloniki.

@@

We have observed ECE cases exceeding the TSI by $400\,\text{W}/\text{m}^{2}$ with
$75\,\%$ of the cases under $200\,\text{W}/\text{m}^{2}$. The climatological
characteristics of the ECE events showed that the most active months are spread in
half of the year and particularly in the periods March -- June and September --
October.  The magnitude of the ECE events identified in Thessaloniki events does not
exceed the values reported for sites with more favourable conditions for the
phenomenon [e.g., @Cordero2023]. Some of the characteristics of CE and ECE events we
analyzed have strong similarities with the results of @Vamvakas2020 for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude of OI.
